Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-03-08 analysis/paper/EDA_figure_1.Rmd
added weather normals figure

>>>>

## Seasonality

Use INMET normals for Manaus from https://portal.inmet.gov.br/normais

```{r}
normals <-
  tibble(
  month = 1:12,
  temp_mean = c(26.30, 26.30, 26.30, 26.40, 26.60, 26.70, 27.00, 27.60, 28.00, 28.00, 27.60, 26.90),
  temp_min  = c(23.10, 23.10, 23.20, 23.20, 23.40, 23.00, 23.10, 23.40, 23.70, 23.90, 23.70, 23.50),
  temp_max  = c(30.90, 30.80, 30.90, 31.00, 31.10, 31.40, 32.10, 33.10, 33.50, 33.40, 32.60, 31.70),
  precip    = c(287.0, 295.1, 300.0, 319.0, 246.9, 118.3, 75.40, 64.30, 76.30, 104.1, 169.2, 245.6)
)

precip <-
  ggplot(normals, aes(x = as.factor(month), y = precip)) + 
  geom_col(fill = "darkblue") +
  scale_y_continuous("Precipitation (mm)", expand = expansion(mult = c(0.05, 0.1))) +
  theme_bw()

temp <- 
  ggplot(normals, aes(x = month)) +
  geom_line(aes(y = temp_mean), color = "red") +
  geom_line(aes(y = temp_min), color = "red", linetype = 2) +
  geom_line(aes(y = temp_max), color = "red", linetype = 2) +
  scale_x_continuous("Month", breaks = 1:12) +
  scale_y_continuous("Temperature (ºC)", expand = expansion(mult = 0.3)) +
  theme_bw()

(precip + theme(axis.title.x = element_blank(), axis.text.x = element_blank()))/
  (temp) + plot_annotation(tag_levels = "a", tag_suffix = ")")
ggsave(here("analysis", "figures", "manaus_normals.png"))
```

### 2021-03-08 analysis/paper/paper.Rmd
added weather normals figure

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
There is a pronounced dry season from June to October **(Figure?)**.

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016].
We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993].
SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017].
The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).

>>>>

```{r}
cap <- "Weather normals for Manaus, Brazil (3º6'S, 60º1'W) for 1981–2010. Precipitation (a) shows a marked dry seasons from June through October. Temperature (b) varies less throughout the year. Mean monthly temperature is shown in the solid red line and monthly minimum and maximum temperatures are shown with the lower and upper dashed lines, respectively. Data from Brazilian National Institute of Meterology (INMET)."
```

```{r normals, fig.cap=cap}
include_graphics(here("analysis", "figures", "manaus_normals.png"))
```

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy [@rankin-de-meronaPreliminaryResultsLargescale1992] and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999].
The soils in the reserves are nutrient-poor xanthic ferrosols; their water retention capacity is poor despite having a high clay content.
Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm.
There is a pronounced dry season from June to October (Figure \@ref(fig:normals)).

@@

Data on precipitation and potential evapotranspiration in our sites were obtained from a published gridded dataset (0.25º x 0.25º resolution) built using data from 3,625 ground-based weather stations across Brazil [@xavierDailyGriddedMeteorological2016]. We used these data to calculate the standardized precipitation evapotranspiration index (SPEI) [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI is a proxy for meteorological drought that integrates precipitation and evapotranspiration anomalies over a specified time scale.
Positive SPEI values for a given month indicate conditions wetter than the historical average for that month, while negative values of SPEI indicate droughts with intensity categorized as mild (0 to -1), moderate (-1 to -1.5), severe (-1.5 to -2), or extreme (\< -2) [@mckeeRelationshipDroughtFrequency1993]. SPEI can be calculated to represent different temporal scales of drought; we used 3-month SPEI because---given its shallow roots and rhizome---*H. acuminata* relies primarily on soil moisture rather than deeper water sources that can persist for longer timescales [@vicente-serranoMultiscalarDroughtIndex2010].
SPEI calculations were done using the `SPEI` package in `r version$version.string` [@rcoreteamLanguageEnvironmentStatistical2020; @begueriaSPEICalculationStandardised2017]. The timing of drought events based on these SPEI calculations is consistent with that resulting from SPEI calculated with other data sources (e.g., nearby weather stations, the global SPEI database [@begueriaSbegueriaSPEIbaseVersion2017], TRMM [@trmm3b43]), though the magnitude of drought can sometimes differ (**Supplemental \_\_**).
<!--# maybe add a sentence emphasizing that SPEI is monthly calculation -->
